# Data Manipulation with Base Functions

We will introduce how to manipulate with different datasets using base functions in R.

## Import and save datasets

There couple of ways to importing and saving different types of datasets [@quickr1;@quickr2].

### Import data from commonly-used file types

**CSV file**
```{r, eval=FALSE}
mydata <- read.csv('c:/minneapolis.csv', ## file location and name
                   header = TRUE, ## read the first 
                   sep = ',') ## which type of separation
```

**EXCEL file**
```{r, eval=FALSE}
library(readxl)
dataset <- read_excel('c:/mydata.xlsx', # file location and name
                      sheet = 'data') # name or index of the sheet
```

**dta STATA file**
```{r, eval=FALSE}
library(foreign)
mydata <- read.dta('c:/mydata.dta') # file location and name
```

Below is an easy way to load datasets in RStudio.

**System tool**
Besides importing data by codes, you could also import data with the system tool. If this is your first time to use this tool, there may be a process to install the packages depending your options. But don't worry, RStudio can do it by itself. You just need to click the button to approve the installation.

```
File -> Import dataset -> choose the type of dataset you want to import
```
There are some other options or parameters you could set in the import functions listed above (e.g., specify a variable type or try to skip some of the rows). Sometimes it is sort of complex to do it. So for me, I just do those things after I import the dataset.

### Save file
**CSV file**
```{r, eval=FALSE}
write.csv(df, # data
          'c:/filename.csv') # file location and name
```

**EXCEL file**
```{r, eval=FALSE}
library(xlsx)
write.xlsx(mydata, # data
           "c:/mydata.xlsx") # file location and name
```

**dta STATA file**
```{r, eval=FALSE}
library(foreign)
write.dta(mydata, "c:/mydata.dta")
```

Usually, it takes less time to save file in CSV and CSV file has a smaller size in storage.

### File location

As you can see in the examples, you need to specify the location (or path) of the file to make sure that R could find your file in the right position. Usually you could find it by checking the system property of the file.

You could avoid this. First, put your R file and dataset in the same folder. Then start the R file by double clicking it. R will use the folder where the R file locates as the working folder. Then you could only specify the name of the file.

This is recommended. Since it will be easier for others to check your codes since they do not need to change the path of the file.

## View data

You could view the variable names and simple description in the ``Environment`` pane on the right-top position of RStudio. If you want to view more information, click the variable name and view the variable in a new window. Here, we use the Minneapolis one-year ACS data as a example. This dataset contains the demographic information of the respondents in the city of Minneapolis in the ACS survey from 2010 to 2019.

```{r}
minneapolis <- read.csv('minneapolis.csv')
```

You could also view the data in the new window by ``View()`` function. Please pay attention that it is initial-capitalized.

```{r, eval=FALSE}
View(minneapolis)
```

View the first ten observations (rows) in the dataset. <dbl> stands for double class, which is a sub-type of numerical variable type.
```{r}
head(minneapolis, n = 10)
```

view the last five observations in the dataset.

```{r}
tail(minneapolis, n = 5)
```


List the variables in the dataset [@quickr3].
```{r}
names(minneapolis)
colnames(minneapolis)
```

List the structure of the dataset.
```{r}
str(minneapolis)
```

The table below lists the variable names and their descriptions.

+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| Variable | Name                   | Description                                                                                                                                |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| YEAR     | Year                   | The year when the data was collected.                                                                                                      |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| SEX      | Gender                 | Gender of the respondent.\                                                                                                                  |
|          |                        | 1 = Male\                                                                                                                                   |
|          |                        | 2 = Female\                                                                                                                                 |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| AGE      | Age                    | Age of the respondent (0 indicates age is smaller than 1).                                                                                 |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| RACE     | Race                   | Race of the respondent.\                                                                                                                    |
|          |                        | 1 = White\                                                                                                                                  |
|          |                        | 2 = Black/African American/Negro\                                                                                                           |
|          |                        | 3 = American Indian or Alaska Native\                                                                                                       |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| HISPAN   | Hispanic origin        | Hispanic origin of the respondent.\                                                                                                         |
|          |                        | 0 = Not Hispanic\                                                                                                                           |
|          |                        | 1 = Mexican\                                                                                                                                |
|          |                        | 2 = Puerto Rican\                                                                                                                           |
|          |                        | 3 = Cuban\                                                                                                                                  |
|          |                        | 4 = Other\                                                                                                                                  |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| EDUC     | Educational attainment | Respondents' educational attainment, as measured by the highest year of school or degree completed.\                                        |
|          |                        | 0 = N/A or no schooling\                                                                                                                    |
|          |                        | 1 = Nursery school to grade 4\                                                                                                              |
|          |                        | 2 = Grade 5, 6, 7, or 8\                                                                                                                    |
|          |                        | 3 = Grade 9\                                                                                                                                |
|          |                        | 4 = Grade 10\                                                                                                                               |
|          |                        | 5 = Grade 11\                                                                                                                               |
|          |                        | 6 = Grade 12\                                                                                                                               |
|          |                        | 7 = 1 year of college\                                                                                                                      |
|          |                        | 8 = 2 years of college\                                                                                                                     |
|          |                        | 9 = 3 years of college\                                                                                                                     |
|          |                        | 10 = 4 years of college\                                                                                                                    |
|          |                        | 11 = 5+ years of college                                                                                                                   |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| EMPSTAT  | Employment status      | Whether the respondent was a part of the labor force -- working or seekingwork -- and, if so, whether the person was currently unemployed.\ |
|          |                        | 0 = N/A\                                                                                                                                    |
|          |                        | 1 = Employed\                                                                                                                               |
|          |                        | 2 = Unemployed\                                                                                                                             |
|          |                        | 3 = Not in labor force                                                                                                                     |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| INCTOT   | Total personal income  | Each respondent's total pre-tax personal income or losses from all sources for the previous year                                            |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+
| FTOTINC  | Total family income    | The total pre-tax money income earned by one's family from all sources for the previous year.                       |
+----------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------+

List the dimensions of the dataset
```{r}
dim(minneapolis)
```

List the number of rows in the dataset
```{r}
nrow(minneapolis)
```

List the number of columns in the dataset.
```{r}
ncol(minneapolis)
```

## Data selection
Select one column with 5 rows of observations.
```{r}
head(minneapolis$YEAR, n = 5) # by name
head(minneapolis[1], n = 5) # by index
```

Select several columns with 5 rows of observations.
```{r}
head(minneapolis[c('YEAR', 'SEX')], n = 5) # by name
head(minneapolis[c(1, 3, 5)], n = 5) # by index
```

Select one row by index
```{r}
minneapolis[2,] # by index number
```

Select several rows
```{r}
minneapolis[2:3, ] # by index
minneapolis[c(1,5,9), ] # by index
```

## Conditional selection

```{r}
newdata <- minneapolis[minneapolis$YEAR == 2010, ]
head(newdata)
```

```{r}
newdata <- minneapolis[(minneapolis$YEAR == 2010) & (minneapolis$EDUC < 5), ]
head(newdata)
```

```{r}
newdata <- minneapolis$INCTOT[minneapolis$YEAR == 2010]
head(newdata)
```

## Deal with missing values
In R, the missing values is presented as ``NA``.

Test the existence of missing values with ``is.na()`` function. We use an revised old example here.

```{r}
ID <- c(1:4) # create variable ID
Name <- c('A', NA, 'C', 'D') # create variable Name
Score <- c(69.5, 77.5, NA, 90) # create variable Score
df <- data.frame(ID, Name, Score) # combine the varibles into one data frame called df
is.na(df)
```

Assign missing values
```{r}
df$Score[df$Score == 90] <- NA
df
```

NAs will influence some functions.
```{r}
mean(df$Score) # get the mean value (does not ignore NA)
mean(df$Score, na.rm=TRUE) # (ignore NA)
```

Test if the observations in the dataset has NAs.
```{r}
complete.cases(df)
```

Find the observations with no NAs.
```{r}
na.omit(df)
```

## Subset

``subset()`` is another way to select the data you want.

Select observations in the Minneapolis population dataset when ``YEAR`` is equal to 2010.
```{r}
newdata <- subset(minneapolis, YEAR == 2010)
head(newdata)
```

Select two columns from the dataset.
```{r}
newdata <- subset(minneapolis, select = c('YEAR', 'AGE'))
head(newdata, n = 5)
```

## Merge two datasets

``merge()`` function does the same work as ``vlookup()`` in excel and 'Join' function in ArcGIS. It links two datasets based on their common variable (the variable they both have).



```{r}
ID <- c(1:4) # create variable ID
Name <- c('A', 'B', 'C', 'D') # create variable Name
Score1 <- c(69.5, 77.5, 99, 90) # create variable Score1
df1 <- data.frame(ID, Name, Score1) # combine the variables into one data frame called df1
df1

Name <- c('A', 'D', 'C') # create variable Name
Score2 <- c(98, 46, 55) # create variable Score2
df2 <- data.frame(Name, Score2) # combine the variables into one data frame called df2
df2
```

```{r}
merge(df1, df2,       # data frames need to be merged
      by = 'Name',    # name of the column/variable used for merging
      all.x = TRUE)   # keep all observations in the first data frame after merging
```

```{r}
merge(df1, df2, 
      by = 'Name',
      all.y = TRUE) # keep all observations in the first dataframe after merging
```

You could keep all the observations in both two data frames by set ``all = TRUE`` in the function. We will cover more about this in our lecture about ``dplyr``.

## Column operation

Column operation or vector operation is a very important idea in R. It applies the operations in two columns or the function in one column directly rather than applies them on each element one by one. 
```{r}
ID <- c(1:4) # create variable ID
Name <- c('A', 'B', 'C', 'D') # create variable Name
Score1 <- c(69.5, 77.5, 99, 90) # create variable Score1
Score2 <- c(98, 46, 55, 70) # create variable Score2
df <- data.frame(ID, Name, Score1, Score2) # combine the varibles into one data frame called df
df
```

Create a new column in the data frame to calculate the sum of score 1 and score 2
```{r}
df$totalsocre <- df$Score1 + df$Score2
df
```

Create a new column in the data frame to calculate the mean of score 1 and score 2
```{r}
df$meansocre <- (df$Score1 + df$Score2)/2
df
```



