# Data Visualization with Base Functions

We go through the base plotting functions in R in this chapter.

```{r}
## import the Minneapolis ACS dataset
minneapolis <- read.csv('minneapolis.csv')
```


## Scatter plot

Scatter plot is a good way to show the distribution of data points.

```{r}
plot(minneapolis$AGE, minneapolis$INCTOT) # with first as x, and second as y
```

Or, you could use the variable names directly and indicate the dataset as the codes below. You will get the same result.

```{r}
plot(INCTOT ~ AGE, data = minneapolis) # you have to specify the name of the data frame here
```


## Line plot

You could transfer the scatter plot above to a line plot by just adding a ``type`` variable to indicate that you are plotting a line. Line plot is good for presenting the trend of a variable changing by time.

```{r}
library(dplyr)
mean_income <- minneapolis %>% 
    group_by(YEAR) %>% 
    summarise(AvgInc = mean(INCTOT, na.rm = T))

plot(AvgInc ~ YEAR,
     data = mean_income,
     type = 'l') ## type indicates the line type with l
```

You could also choose another type by changing the value of ``type``, as the one below.

```{r}
plot(AvgInc ~ YEAR,
     data = mean_income,
     type = 'b') # b for both line and pint
```

You could use ``help(plot)`` to check more styles of the plots.

## Bar plot

Bar plot is a good way to compare the values in each year or for each item. You could use ``barplot()`` to draw it.

```{r}
barplot(mean_income$AvgInc,
        names.arg = mean_income$YEAR) # names.org indicates the vector of names to be plotted under each bar
```


## Add more elements in the plots

For a reader-friendly plots, you have to add more information such as title, labels, and legend. For the plot above, we could use the codes below to make it more informative.

```{r}
barplot(mean_income$AvgInc,
        names.arg = mean_income$YEAR,
        main = 'Bar plot of the average personal income in Minneapolis (2010-2019)', # add title for the plot
        xlab = 'Year', # add label tag for the x-axis
        ylab = 'Personal income (dollars)',  # add label for the y-axis
        ylim = c(0, 65000), # set the range of y axis, you could set the range of x axis with xlim
        legend = 'Personal income') # add legend name
```

## Pie chart

Pie chart is a good way to show the share of each part. You could use ``pie()`` function to draw a pie chart in R.

```{r}
minneaplis_race <- minneapolis %>% 
      mutate(
    RACE = case_when( ## change RACE from numeric values to racial categories
      RACE == 1 ~ 'White',
      RACE == 2 ~ 'African American',
      RACE == 3 ~ 'Other'
    )
  ) %>% 
    group_by(RACE) %>% 
    summarise(count = n())


pie(minneaplis_race$count, # value for each piece
    labels = minneaplis_race$RACE) # label for each piece
```

## Boxplot

Box plot is also called box-whisker plot. It is to present the distribution of the dataset based on their quartiles. In R, you could use ``boxplot()`` to draw a box plot.

```{r}
t <- c(1, 5, 10, 7, 8, 10, 11, 19)
boxplot(t, range = 0) # set range = 0 makes the whiskers reach the smallest and largest values in the dataset
```

```{r}
t <- c(1, 5, 10, 7, 8, 10, 11, 19)
boxplot(t, range = 1) # set range = 1 makes the the whiskers extend to the most extreme data point which is no more than range times the interquartile range from the box
```

## Color in R

You could change the color of the plots by adding ``col = `` in the functions. For example.

```{r}
plot(AvgInc ~ YEAR,
     data = mean_income,
     type = 'b',
     col = 'YellowGreen') # specify the name of the color
```

Here is a [link](https://html-color-codes.info/color-names/) where you could find the name of the color.

You could also use the hexadecimal color code to indicate the color. For example.

```{r}
barplot(mean_income$AvgInc,
        names.arg = mean_income$YEAR,
        main = 'Bar plot of the average personal income in Minneapolis (2010-2019)',
        xlab = 'Year', 
        ylab = 'Personal income (dollars)',
        ylim = c(0, 65000),
        legend = 'Personal income',
        col = '#009999') # use the hexadecimal color code, you need to start it with the hash tag
```

By the same [link](https://html-color-codes.info/color-names/), you could also find the hexadcimal code for each color.
